A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-09 podc_paper/correctness-2.tex
minor edits

\begin{customlemma}{\ref{lem::doubleRefresh}$'$}\label{lem::doubleRefreshGC}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}($v$) by the same process.
All operations contained $v$'s children when $R_1$ begins
are contained in $v$ when $R_2$ terminates.
\end{customlemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{casGC} or terminates at line \ref{addOPGC}, the claim follows
from \Cref{successfulRefreshGC}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{casGC}.
\here{To be continued....}

@@

>>>>

\begin{customlemma}{\ref{lem::doubleRefresh}$'$}\label{lem::doubleRefreshGC}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}($v$) by the same process.
All operations contained $v$'s children when $R_1$ begins
are contained in $v$ when $R_2$ terminates.
\end{customlemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{casGC} or terminates at line \ref{addOPGC}, the claim follows
from \Cref{successfulRefreshGC}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{casGC}.
Then some other \op{CAS} on $v.\fld{blocks}$ succeeds between the time each \op{Refresh}
reads $v.\fld{blocks}$ at line \ref{refreshReadTGC} and performs its \op{CAS} at line \ref{casGC}.
Consider the \op{Refresh} $R_3$ that does this successful \op{CAS} during $R_2$.
$R_3$ must have read $v.\fld{blocks}$ after the successful \op{CAS} during $R_1$.
The claim follows from \Cref{successfulRefreshGC} applied to $R_3$.
\end{proof}

\Cref{lem::appendExactlyOnce} can then be proved in the same way as in Section \ref{sec::propagating}.
We modified the way \var{super} fields are computed, so we prove an analogue of \Cref{superRelation} to show that \fld{super} fields are (nearly) accurate.

\begin{customlemma}{$\ref{superRelation}'$}\label{superRelationGC}
Let $B=v.\var{blocks}[b]$.
  If $v.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
\end{customlemma}
\begin{proof}

@@

The correctness of \op{GetEnqueue} and \op{indexDequeue}, which are identical to the original version of the implementation are dependent only on the fact that we do not discard blocks needed by those routines.
The following lemmas show this.

\here{prove some lemmas about not throwing away useful information, then restate correctness lemmas about getenq and indexdeq that say they work correctly OR the operation is already propagated/complete.}

### 2023-01-09 podc_paper/correctness.tex
working on GC section
minor edits

\Eric{I think this case was missing from the proof in the thesis}
Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
Intuitively, since there are no operations in $v$'s children to promote, $op$ is already in $v$.
We formalize this intuition.
The value computed at line \ref{createSumEnq} is
\begin{eqnarray*}
\var{num\sub{enq}} 
&=& v.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + v.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} - v.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
&=& v.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + v.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} \\
&&\mbox{ }- v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} - v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}
\end{eqnarray*}
It follows from \Cref{lem::sum} that $num\sub{enq}$ is the total number of enqueues contained in the blocks
$v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
$v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright+1..new.\eright]$.
Similarly, $num\sub{deq}$ is the total number of dequeues contained in these blocks.
Since $num\sub{enq}+num\sub{deq}=0$ at line \ref{testEmpty},
these blocks contain no operations.
By \Cref{blockNotEmpty}, this means the ranges of blocks are empty, so that $v.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft \geq i$.
Hence, $B$ is already a subblock of some block in $v$, so $v$ contains $op$.
\end{proof}

@@

\begin{lemma}\label{lem::doubleRefresh}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}($v$) by the same process.
When $R_2$ terminates, $v$ contains all operations that $v$'s children contained when $R_1$ was invoked.
\end{lemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{cas} or terminates at line \ref{addOP}, the claim follows
from \Cref{successfulRefresh}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{cas}.
Let $h_1$ and $h_2$ be the values $R_1$ and $R_2$ read from $v.\head$ at line \ref{readHead}.
By \Cref{lem::headInc}, $h_2>h_1$.
By \Cref{lem::headProgress}, $v.blocks[h_2]=\nl$ when $R_1$ executes line \ref{readHead}.
Since $R_2$ fails its \op{CAS} on $v.blocks[h_2]$, some other \op{Refresh} $R_3$ must have done
a successful \op{CAS} on $v.blocks[h_2]$ before $R_2$'s \op{CAS}.
$R_3$ must have executed line \ref{readHead} after $R_1$, since $R_3$ read the value $h_2$ from $v.\head$ and the value of $v.\head$ is non-decreasing, by \Cref{nonDecreasingHead}.
Thus, all operations contained in $v$'s children when $R_1$ begins
are also contained in $v$'s children when $R_3$ later executes line \ref{readHead}.
By \Cref{successfulRefresh}, these operations are contained in $v$ when $R_3$ performs its successful \op{CAS},
which is before $R_2$ terminates.
\end{proof}

@@

Next, we show that $B.\fld{super}\geq s-1$.
The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $v.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ values are always at least 1, so $B.\fld{super} \geq 1$.
So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the remainder of the proof.
By \Cref{lem::headProgress}, $v.\fld{parent.blocks}[s-1]\neq \nl$.  Let $R_{s-1}$ be the call to
$\op{Refresh}(v.\fld{parent})$ that installed the block in $v.\fld{parent.blocks}[s-1]$.
Let $r'$ be the step when $R_{s-1}$ reads $s-1$ in $v.\fld{parent}.\head$ at line \ref{readHead}.
This read $r'$ must be before $B$ is installed in $v$;
otherwise, \Cref{successfulRefresh} would imply that $B$ is a subblock of one of 
$v.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
Now, consider the call to \op{Advance}($v, b$) that writes $B.\fld{super}$.
It is invoked either 
at line \ref{helpAdvance} after seeing $v.\fld{blocks}[h]\neq \nl$ at line \ref{ifHeadnotNull}
or at line \ref{advance} after ensuring $v.\fld{blocks}[h]\neq \nl$ at line~\ref{cas}.
Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.
By \Cref{nonDecreasingHead}, $v.\fld{parent}.\head$ is non-decreasing, so 
the value this \op{Advance} reads in $v.\fld{parent}.\head$ and
writes in $B.\fld{super}$ is greater than or equal to the value $s-1$ that $r'$ reads in $v.\fld{parent}.\head$.
\end{proof}

>>>>

%\Eric{I think this case was missing from the proof in the thesis}
Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
Intuitively, since there are no operations in $v$'s children to promote, $op$ is already in $v$.
We formalize this intuition.
The value computed at line \ref{createSumEnq} is
\begin{eqnarray*}
\var{num\sub{enq}} 
&=& v.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + v.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} - v.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
&=& v.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + v.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} \\
&&\mbox{ }- v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} - v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}
\end{eqnarray*}
It follows from \Cref{lem::sum} that $num\sub{enq}$ is the total number of enqueues contained in the blocks
$v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
$v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright+1..new.\eright]$.
Similarly, $num\sub{deq}$ is the total number of dequeues contained in these blocks.
Since $num\sub{enq}+num\sub{deq}=0$ at line \ref{testEmpty},
these blocks contain no operations.
By \Cref{blockNotEmpty}, this means the ranges of blocks are empty, so that $v.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft \geq i$.
Hence, $B$ is already a subblock of some block in $v$, so $v$ contains $op$.
\end{proof}

@@

\begin{lemma}\label{lem::doubleRefresh}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}($v$) by the same process.
All operations contained $v$'s children when $R_1$ begins
are contained in $v$ when $R_2$ terminates.
\end{lemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{cas} or terminates at line \ref{addOP}, the claim follows
from \Cref{successfulRefresh}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{cas}.
Let $h_1$ and $h_2$ be the values $R_1$ and $R_2$ read from $v.\head$ at line \ref{readHead}.
By \Cref{lem::headInc}, $h_2>h_1$.
By \Cref{lem::headProgress}, $v.blocks[h_2]=\nl$ when $R_1$ executes line \ref{readHead}.
Since $R_2$ fails its \op{CAS} on $v.blocks[h_2]$, some other \op{Refresh} $R_3$ must have done
a successful \op{CAS} on $v.blocks[h_2]$ before $R_2$'s \op{CAS}.
$R_3$ must have executed line \ref{readHead} after $R_1$, since $R_3$ read the value $h_2$ from $v.\head$ and the value of $v.\head$ is non-decreasing, by \Cref{nonDecreasingHead}.
Thus, all operations contained in $v$'s children when $R_1$ begins
are also contained in $v$'s children when $R_3$ later executes line \ref{readHead}.
By \Cref{successfulRefresh}, these operations are contained in $v$ when $R_3$ performs its successful \op{CAS},
which is before $R_2$ terminates.
\end{proof}

@@

Next, we show that $B.\fld{super}\geq s-1$.
The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $v.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ values are always at least 1, so $B.\fld{super} \geq 1$.
So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the remainder of the proof.
By \Cref{lem::headProgress}, $v.\fld{parent.blocks}[s-1]\neq \nl$.  Let $R_{s-1}$ be the call to
$\op{Refresh}(v.\fld{parent})$ that installed the block in $v.\fld{parent.blocks}[s-1]$.
Let $r'$ be the step when $R_{s-1}$ reads $s-1$ in $v.\fld{parent}.\head$ at line \ref{readHead}.
This read $r'$ must be before $B$ is installed in $v$;
otherwise, \Cref{successfulRefresh} would imply that $B$ is a subblock of one of 
$v.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
Now, consider the call to \op{Advance}($v, b$) that writes $B.\fld{super}$.
It is invoked either 
at line \ref{helpAdvance} after seeing $v.\fld{blocks}[b]\neq \nl$ at line \ref{ifHeadnotNull}
or at line \ref{advance} after ensuring $v.\fld{blocks}[b]\neq \nl$ at line~\ref{cas}.
Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.
By \Cref{nonDecreasingHead}, $v.\fld{parent}.\head$ is non-decreasing, so 
the value this \op{Advance} reads in $v.\fld{parent}.\head$ and
writes in $B.\fld{super}$ is greater than or equal to the value $s-1$ that $r'$ reads in $v.\fld{parent}.\head$.
\end{proof}
